---
title: "arrow"
author: "Romain François"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{arrow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(arrow, warn.conflicts = FALSE)
```

# Install arrow

## CRAN 

`arrow` is not yet available from CRAN

## Installing from source

The `arrow` package is currently only available from github. To install the development version, you first need to 
install the C++ library so that `pkg-config` finds it: 

```shell
git clone https://github.com/apache/arrow.git
cd arrow/cpp && mkdir release && cd release

# It is important to statically link to boost libraries
cmake .. -DCMAKE_BUILD_TYPE=Release -DARROW_BOOST_USE_SHARED:BOOL=Off
make install
```

Then you can install the R package with [remotes::install_github()](https://remotes.r-lib.org/reference/install_github.html)

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("apache/arrow/r")
```

This will install the correct versions of packages `arrow` depend on. 

# Development

## System Requirements

## Environment Setup and Build

## Build and test

## Developing on Windows

# Memory and IO Interfaces

# Data Type and In-Memory Data Model

Apache Arrow defines columnar array data structures by composing type metadata with memory buffers. These data 
structures are exposed in R as a set of interrelated [R6](https://r6.r-lib.org) classes. 

 - Type metadata: Instances of `arrow::DataType`, which describe a logical array type
 - Schemas: Instances of `arrow::Schema`, which describes a named collection of types. These
   can be thought of as the column types in a table-like object. 
 - Arrays: Instances of `arrow::Array`, which are atomic, contiguous columnar data structures 
   composed from `arrow::Buffer` objects. 
 - Record Batches: Instances of `arrow::RecordBatch` which are a collection of `Array` objects 
   with a particular schema
 - Tables: Instances of `arrow::Table`, a logical table data structures in which each column
   consists of one or more `Array` objects of the same logical type. 
   
## Type metadata

Apache Arrow defines language agnostic column-oriented data structures for array data. These include: 

 - Fixed-length primitive types: numbers, booleans, data and times, fixed size binary and other values
   that fit into a given number
 - Variable length primitive types: binary, string
 - Nested types: list, struct and union
 - Dictionary type: An encoded categorical type
 
Each logical type in `arrow` has a corresponding factory function for creating an instance of that type object in R. 

```{r}
library(arrow, warn.conflicts = FALSE)
t1 <- int32()
t2 <- utf8()
t5 <- timestamp(TimeUnit$MILLI)

t1
t2
t5
```

We use the name *logical* type because the *physical* storage may be the same for one or more types. For
example `int64`, `float64` and `timestamp[ms]` all occupy 64 bites per value. 

These objects are *metadata*, they are used for describing the data in arrays, schemas and record batches. 

