---
title: "Data analysis with dplyr syntax"
description: >
  Learn how to use the dplyr backend supplied by arrow 
output: rmarkdown::html_vignette
---

The arrow package provides functionality allowing users to manipulate tabular Arrow data (`Table` and `Dataset` objects) with familiar `{dplyr}` syntax. To enable this functionality, ensure that the arrow and dplyr packages are both loaded. In this article we will take the `starwars` data set included in dplyr, convert it to an Arrow Table, and then analyze this data. Note that, although these examples all use an in-memory `Table` object, the same functionality works for an on-disk `Dataset` object with only minor differences in behavior (documented later in the article).

To get started let's load the packages and create the data:

```{r}
library(dplyr, warn.conflicts = FALSE)
library(arrow, warn.conflicts = FALSE)

sw <- arrow_table(starwars, as_data_frame = FALSE)
```

## One-table dplyr verbs

The arrow package provides support for the dplyr one-table verbs, allowing users to construct data analysis pipelines in a familiar way. The example below shows the use of `filter()`, `rename()`, `mutate()`, `arrange()` and `select()`:

```{r}
result <- sw %>%
  filter(homeworld == "Tatooine") %>%
  rename(height_cm = height, mass_kg = mass) %>%
  mutate(height_in = height_cm / 2.54, mass_lbs = mass_kg * 2.2046) %>%
  arrange(desc(birth_year)) %>%
  select(name, height_in, mass_lbs)
```

It is important to note that arrow uses lazy evaluation to delay computation until the result is explicitly requested. This speeds up processing by enabling the Arrow C++ library to perform multiple computations in one operation. As a consequence of this design choice, we have not yet performed computations on the `sw` data. The `result` variable is an object with class `arrow_dplyr_query` that represents all the computations to be performed:

```{r}
result
```

To perform these computations and materialize the result, we call
`compute()` or `collect()`. The difference between the two determines what kind of object will be returned. Calling `compute()` returns an Arrow Table, suitable for passing to other arrow or dplyr functions:

```{r}
compute(result)
```

In contrast, `collect()` returns an R data frame, suitable for viewing or passing to other R functions for analysis or visualization:

```{r}
collect(result)
```

The arrow package has broad support for single-table dplyr verbs, including those that compute aggregates. For example, it supports `group_by()` and `summarize()`, as well as commonly-used convenience functions such as `count()`:

```{r}
sw %>%
  group_by(species) %>%
  summarize(mean_height = mean(height, na.rm = TRUE)) %>%
  collect()

sw %>%
  count(gender) %>%
  collect()
```

Note, however, that window functions such as `ntile()` are not yet supported. 

## Two-table dplyr verbs

Equality joins (e.g. `left_join()`, `inner_join()`) are supported for joining multiple tables. This is illustrated below:

```{r}
jedi <- data.frame(
  name = c("C-3PO", "Luke Skywalker", "Obi-Wan Kenobi"),
  jedi = c(FALSE, TRUE, TRUE)
)

sw %>%
  select(1:3) %>%
  right_join(jedi) %>%
  collect()
```

## Expressions within dplyr verbs

Inside dplyr verbs, Arrow offers support for many functions and operators, with common functions mapped to their base R and tidyverse equivalents: you can find a [list of supported functions within dplyr queries](../reference/acero.html) in the function documentation. If there are additional functions you would like to see implemented, please file an issue as described in the [Getting help](https://arrow.apache.org/docs/r/#getting-help) guidelines.

## Registering custom bindings

The arrow package makes it possible for users to supply bindings for custom functions in some situations using `register_scalar_function()`. To operate correctly, the to-be-registered function must have `context` as its first argument, as required by the query engine. For example, suppose we wanted to implement a function that converts a string to snake case (a greatly simplified version of `janitor::make_clean_names()`). The function could be written as follows:

```{r}
to_snake_name <- function(context, string) {
  replace <- c(`'` = "", `"` = "", `-` = "", `\\.` = "_", ` ` = "_")
  string %>%
    stringr::str_replace_all(replace) %>%
    stringr::str_to_lower() %>%
    stringi::stri_trans_general(id = "Latin-ASCII")
}
```

To call this within an arrow/dplyr pipeline, it needs to be registered:

```{r}
register_scalar_function(
  name = "to_snake_name",
  fun = to_snake_name,
  in_type = utf8(),
  out_type = utf8(),
  auto_convert = TRUE
)
```

In this expression, the `name` argument specifies the name by which it will be recognized in the context of the arrow/dplyr pipeline and `fun` is the function itself. The `in_type` and `out_type` arguments are used to specify the expected data type for the input and output, and `auto_convert` specifies whether arrow should automatically convert any R inputs to their Arrow equivalents. 

Once registered, the following works:

```{r}
sw %>%
  mutate(name, snake_name = to_snake_name(name), .keep = "none") %>%
  collect()
```

To learn more, see `help("register_scalar_function", package = "arrow")`. 

## Handling unsupported expressions

For dplyr queries on Table objects, which are held in memory and should
usually be representable as data frames, if the arrow package detects
an unimplemented function within a dplyr verb, it automatically calls
`collect()` to return the data as an R data frame before processing
that dplyr verb. As an example, neither `lm()` nor `residuals()` are 
implemented, so if we write code that computes the residuals for a
linear regression model, this automatic collection takes place:

```{r}
sw %>%
  filter(!is.na(height), !is.na(mass)) %>%
  transmute(name, height, mass, res = residuals(lm(mass ~ height)))
```

For queries on `Dataset` objects -- which can be larger
than memory -- arrow is more conservative and always raises an
error if it detects an unsupported expression. To illustrate this 
behavior, we can write the `starwars` data to disk and then open
it as a Dataset. When we use the same pipeline on the Dataset, 
we obtain an error:

```{r, error=TRUE}
# write and open starwars dataset
dataset_path <- tempfile()
write_dataset(starwars, dataset_path)
sw2 <- open_dataset(dataset_path)

# dplyr pipeline with unsupported expressions
sw2 %>%
  filter(!is.na(height), !is.na(mass)) %>%
  transmute(name, height, mass, res = residuals(lm(mass ~ height)))
```

Calling `collect()` in the middle of the pipeline fixes the issue:

```{r}
sw2 %>%
  filter(!is.na(height), !is.na(mass)) %>%
  collect() %>%
  transmute(name, height, mass, res = residuals(lm(mass ~ height)))
```

## Further reading

- To learn more about multi-file datasets, see the [dataset article](./dataset.html).
- To learn more about user-registered functions, see `help("register_scalar_function", package = "arrow")`.
- To learn more about writing dplyr bindings as an arrow developer, see the [article on writing bindings](./developers/writing_bindings.html). 
