---
title: "Working with Arrow Datasets and dplyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with Arrow Datasets and dplyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Apache Arrow lets you work efficiently with large, multi-file datasets.
The arrow R package provides a [dplyr](https://dplyr.tidyverse.org/) interface to Arrow Datasets,
and other tools for interactive exploration of Arrow data.

This vignette introduces Datasets and shows how to use dplyr to analyze them.

## Example: NYC taxi data

The [New York City taxi trip record data](https://www1.nyc.gov/site/tlc/about/tlc-trip-record-data.page)
is widely used in big data exercises and competitions.
For demonstration purposes, we have hosted a Parquet-formatted version
of about ten years of the trip data in a public Amazon S3 bucket.

The total file size is around 37 gigabytes, even in the efficient Parquet file
format. That's bigger than memory on most people's computers, so you can't just
read it all in and stack it into a single data frame.

In Windows and macOS binary packages, S3 support is included.
On Linux, when installing from source, S3 support is not enabled by default,
and it has additional system requirements.
See `vignette("install", package = "arrow")` for details.
To see if your arrow installation has S3 support, run:

```{r}
arrow::arrow_with_s3()
```

Even with S3 support enabled, network speed will be a bottleneck unless your
machine is located in the same AWS region as the data. So, for this vignette,
we assume that the NYC taxi dataset has been downloaded locally in an "nyc-taxi"
directory.

### Retrieving data from a public Amazon S3 bucket

If your arrow build has S3 support, you can sync the data locally with:

```{r, eval = FALSE}
arrow::copy_files("s3://voltrondata-labs-datasets/nyc-taxi", "nyc-taxi")
# Alternatively, with GCS:
arrow::copy_files("gs://voltrondata-labs-datasets/nyc-taxi", "nyc-taxi")
```

If your arrow build doesn't have S3 support, you can download the files
with the additional code shown below.  Since these are large files, 
you may need to increase R's download timeout from the default of 60 seconds, e.g.
`options(timeout = 300)`.

```{r, eval = FALSE}
bucket <- "https://voltrondata-labs-datasets.s3.us-east-2.amazonaws.com"
for (year in 2009:2022) {
  if (year == 2022) {
    # We only have through Feb 2022 there
    months <- 1:2
  } else {
    months <- 1:12
  }
  for (month in months) {
    dataset_path <- file.path("nyc-taxi", paste0("year=", year), paste0("month=", month))
    dir.create(dataset_path, recursive = TRUE)
    try(download.file(
      paste(bucket, dataset_path, "part-0.parquet", sep = "/"),
      file.path(dataset_path, "part-0.parquet"),
      mode = "wb"
    ), silent = TRUE)
  }
}
```

Note that these download steps in the vignette are not executed: if you want to run
with live data, you'll have to do it yourself separately.
Given the size, if you're running this locally and don't have a fast connection,
feel free to grab only a year or two of data.

If you don't have the taxi data downloaded, the vignette will still run and will
yield previously cached output for reference. To be explicit about which version
is running, let's check whether you're running with live data:

```{r}
dir.exists("nyc-taxi")
```

## Opening the dataset

Because dplyr is not necessary for many Arrow workflows,
it is an optional (`Suggests`) dependency. So, to work with Datasets,
you need to load both arrow and dplyr.

```{r}
library(arrow, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)
```

The first step is to create a Dataset object, pointing at the directory of data.

```{r, eval = file.exists("nyc-taxi")}
ds <- open_dataset("nyc-taxi")
```

The file format for `open_dataset()` is controlled by the `format` parameter, 
which has a default value of `"parquet"`.  If you had a directory
of Arrow format files, you could instead specify `format = "arrow"` in the call.

Other supported formats include: 

* `"feather"` or `"ipc"` (aliases for `"arrow"`, as Feather v2 is the Arrow file format)
* `"csv"` (comma-delimited files) and `"tsv"` (tab-delimited files)
* `"text"` (generic text-delimited files - use the `delimiter` argument to specify which to use)

For text files, you can pass the following parsing options to `open_dataset()`:

* `delim`
* `quote`
* `escape_double`
* `escape_backslash`
* `skip_empty_rows`

For more information on the usage of these parameters, see `?read_delim_arrow()`.

`open_dataset()` was able to automatically infer column values for `year` and `month`
--which are not present in the data files--based on the directory structure. The 
[Hive](https://hive.apache.org/)-style partitioning structure is self-describing, with file paths like

```
year=2009/month=1/part-0.parquet
year=2009/month=2/part-0.parquet
...
```

But sometimes the directory partitioning isn't self describing; that is, it doesn't
contain field names. For example, if instead we had file paths like

```
2009/01/part-0.parquet
2009/02/part-0.parquet
...
```

then `open_dataset()` would need some hints as to how to use the file paths. In this
case, you could provide `c("year", "month")` to the `partitioning` argument, 
saying that the first path segment gives the value for `year`, and the second 
segment is `month`. Every row in `2009/01/part-0.parquet` has a value of 2009 for `year`
and 1 for `month`, even though those columns may not be present in the file.

In either case, when you look at the dataset, you can see that in addition to the columns present
in every file, there are also columns `year` and `month` even though they are not present in the files themselves.

```{r, eval = file.exists("nyc-taxi")}
ds
```
```{r, echo = FALSE, eval = !file.exists("nyc-taxi")}
cat("
FileSystemDataset with 158 Parquet files
vendor_name: string
pickup_datetime: timestamp[ms]
dropoff_datetime: timestamp[ms]
passenger_count: int64
trip_distance: double
pickup_longitude: double
pickup_latitude: double
rate_code: string
store_and_fwd: string
dropoff_longitude: double
dropoff_latitude: double
payment_type: string
fare_amount: double
extra: double
mta_tax: double
tip_amount: double
tolls_amount: double
total_amount: double
improvement_surcharge: double
congestion_surcharge: double
pickup_location_id: int64
dropoff_location_id: int64
year: int32
month: int32
")
```

## Querying the dataset

Up to this point, you haven't loaded any data. You've walked directories to find
files, you've parsed file paths to identify partitions, and you've read the
headers of the Parquet files to inspect their schemas so that you can make sure
they all are as expected.

In the current release, arrow supports the dplyr verbs:

 * `mutate()` and `transmute()`,
 * `select()`, `rename()`, and `relocate()`,
 * `filter()`,
 * `arrange()`,
 * `union()` and `union_all()`,
 * `left_join()`, `right_join()`, `full_join()`, `inner_join()`, and `anti_join()`,
 * `group_by()` and `summarise()`.

At any point in a chain, you can use `collect()` to pull the selected subset of
the data into an in-memory R data frame. 

Suppose you attempt to call unsupported dplyr verbs or unimplemented functions
in your query on an Arrow Dataset. In that case, the arrow package raises an error. However,
for dplyr queries on Arrow Table objects (which are already in memory), the
package automatically calls `collect()` before processing that dplyr verb.

Here's an example: suppose that you are curious about tipping behavior among the
longest taxi rides. Let's find the median tip percentage for rides with
fares greater than $100 in 2015, broken down by the number of passengers:

```{r, eval = file.exists("nyc-taxi")}
system.time(ds %>%
  filter(total_amount > 100, year == 2015) %>%
  select(tip_amount, total_amount, passenger_count) %>%
  mutate(tip_pct = 100 * tip_amount / total_amount) %>%
  group_by(passenger_count) %>%
  summarise(
    median_tip_pct = median(tip_pct),
    n = n()
  ) %>%
  collect() %>%
  print())
```

```{r, echo = FALSE, eval = !file.exists("nyc-taxi")}
cat("
# A tibble: 10 x 3
   passenger_count median_tip_pct      n
             <int>          <dbl>  <int>
 1               1           16.6 143087
 2               2           16.2  34418
 3               5           16.7   5806
 4               4           11.4   4771
 5               6           16.7   3338
 6               3           14.6   8922
 7               0           10.1    380
 8               8           16.7     32
 9               9           16.7     42
10               7           16.7     11

   user  system elapsed
  4.436   1.012   1.402
")
```

You've just selected a subset out of a dataset with around 2 billion rows, computed
a new column, and aggregated it in a few seconds on a modern laptop. How does
this work?

First, the dplyr verbs on the dataset record their actions but don't evaluate on
the data until you run `collect()`.

```{r, eval = file.exists("nyc-taxi")}
ds %>%
  filter(total_amount > 100, year == 2015) %>%
  select(tip_amount, total_amount, passenger_count) %>%
  mutate(tip_pct = 100 * tip_amount / total_amount) %>%
  group_by(passenger_count) %>%
  summarise(
    median_tip_pct = median(tip_pct),
    n = n()
  )
```

```{r, echo = FALSE, eval = !file.exists("nyc-taxi")}
cat("
FileSystemDataset (query)
passenger_count: int64
median_tip_pct: double
n: int32

See $.data for the source Arrow object
")
```

This code returns an output instantly and shows the manipulations you've made, without
loading data from the files. Because the evaluation of these queries is deferred,
you can build up a query that selects down to a small subset without generating
intermediate datasets that would potentially be large.

Second, all work is pushed down to the individual data files,
and depending on the file format, chunks of data within the files. As a result,
you can select a subset of data from a much larger dataset by collecting the
smaller slices from each file—you don't have to load the whole dataset in 
memory to slice from it.

Third, because of partitioning, you can ignore some files entirely.
In this example, by filtering `year == 2015`, all files corresponding to other years
are immediately excluded: you don't have to load them in order to find that no
rows match the filter. Relatedly, since Parquet files contain row groups with
statistics on the data within, there may be entire chunks of data you can
avoid scanning because they have no rows where `total_amount > 100`.

### Processing data in batches

Sometimes you want to run R code on the entire dataset, but that dataset is much
larger than memory. You can use `map_batches` on a dataset query to process
it batch-by-batch.

**Note**: `map_batches` is experimental and not recommended for production use.

As an example, to randomly sample a dataset, use `map_batches` to sample a 
percentage of rows from each batch:

```{r, eval = file.exists("nyc-taxi")}
sampled_data <- ds %>%
  filter(year == 2015) %>%
  select(tip_amount, total_amount, passenger_count) %>%
  map_batches(~ as_record_batch(sample_frac(as.data.frame(.), 1e-4))) %>%
  mutate(tip_pct = tip_amount / total_amount) %>%
  collect()

str(sampled_data)
```

```{r, echo = FALSE, eval = !file.exists("nyc-taxi")}
cat("
tibble [10,918 × 4] (S3: tbl_df/tbl/data.frame)
 $ tip_amount     : num [1:10918] 3 0 4 1 1 6 0 1.35 0 5.9 ...
 $ total_amount   : num [1:10918] 18.8 13.3 20.3 15.8 13.3 ...
 $ passenger_count: int [1:10918] 3 2 1 1 1 1 1 1 1 3 ...
 $ tip_pct        : num [1:10918] 0.1596 0 0.197 0.0633 0.0752 ...
")
```

This function can also be used to aggregate summary statistics over a dataset by
computing partial results for each batch and then aggregating those partial
results. Extending the example above, you could fit a model to the sample data
and then use `map_batches` to compute the MSE on the full dataset.

```{r, eval = file.exists("nyc-taxi")}
model <- lm(tip_pct ~ total_amount + passenger_count, data = sampled_data)

ds %>%
  filter(year == 2015) %>%
  select(tip_amount, total_amount, passenger_count) %>%
  mutate(tip_pct = tip_amount / total_amount) %>%
  map_batches(function(batch) {
    batch %>%
      as.data.frame() %>%
      mutate(pred_tip_pct = predict(model, newdata = .)) %>%
      filter(!is.nan(tip_pct)) %>%
      summarize(sse_partial = sum((pred_tip_pct - tip_pct)^2), n_partial = n()) %>%
      as_record_batch()
  }) %>%
  summarize(mse = sum(sse_partial) / sum(n_partial)) %>%
  pull(mse)
```

```{r, echo = FALSE, eval = !file.exists("nyc-taxi")}
cat("
[1] 0.1304284
")
```

## More dataset options

There are a few ways you can control the Dataset creation to adapt to special use cases.

### Work with files in a directory

If you are working with a single file or a set of files that are not all in the 
same directory, you can provide a file path or a vector of multiple file paths 
to `open_dataset()`. This is useful if, for example, you have a single CSV file 
that is too big to read into memory. You could pass the file path to 
`open_dataset()`, use `group_by()` to partition the Dataset into manageable chunks, 
then use `write_dataset()` to write each chunk to a separate Parquet file—all 
without needing to read the full CSV file into R.

### Explicitly declare column names and data types

You can specify the `schema` argument to `open_dataset()` to declare the columns 
and their data types. This is useful if you have data files that have different 
storage schema (for example, a column could be `int32` in one and `int8` in 
another) and you want to ensure that the resulting Dataset has a specific type.

To be clear, it's not necessary to specify a schema, even in this example of 
mixed integer types, because the Dataset constructor will reconcile differences
like these. The schema specification just lets you declare what you want the 
result to be.

### Explicitly declare partition format

Similarly, you can provide a Schema in the `partitioning` argument of `open_dataset()`
in order to declare the types of the virtual columns that define the partitions.
This would be useful, in the taxi dataset example, if you wanted to keep
`month` as a string instead of an integer.

### Work with multiple data sources

Another feature of Datasets is that they can be composed of multiple data sources.
That is, you may have a directory of partitioned Parquet files in one location,
and in another directory, files that haven't been partitioned.
Or, you could point to an S3 bucket of Parquet data and a directory
of CSVs on the local file system and query them together as a single dataset.
To create a multi-source dataset, provide a list of datasets to `open_dataset()`
instead of a file path, or simply concatenate them like `big_dataset <- c(ds1, ds2)`.

## Writing datasets

As you can see, querying a large dataset can be made quite fast by storage in an
efficient binary columnar format like Parquet or Feather and partitioning based on
columns commonly used for filtering. However, data isn't always stored that way.
Sometimes you might start with one giant CSV. The first step in analyzing data 
is cleaning is up and reshaping it into a more usable form.

The `write_dataset()` function allows you to take a Dataset or another tabular 
data object—an Arrow Table or RecordBatch, or an R data frame—and write
it to a different file format, partitioned into multiple files.

Assume that you have a version of the NYC Taxi data as CSV:

```r
ds <- open_dataset("nyc-taxi/csv/", format = "csv")
```

You can write it to a new location and translate the files to the Feather format
by calling `write_dataset()` on it:

```r
write_dataset(ds, "nyc-taxi/feather", format = "feather")
```

Next, let's imagine that the `payment_type` column is something you often filter
on, so you want to partition the data by that variable. By doing so you ensure
that a filter like `payment_type == "Cash"` will touch only a subset of files
where `payment_type` is always `"Cash"`.

One natural way to express the columns you want to partition on is to use the
`group_by()` method:

```r
ds %>%
  group_by(payment_type) %>%
  write_dataset("nyc-taxi/feather", format = "feather")
```

This will write files to a directory tree that looks like this:

```r
system("tree nyc-taxi/feather")
```

```
## feather
## ├── payment_type=1
## │   └── part-18.feather
## ├── payment_type=2
## │   └── part-19.feather
## ...
## └── payment_type=UNK
##     └── part-17.feather
##
## 18 directories, 23 files
```

Note that the directory names are `payment_type=Cash` and similar:
this is the Hive-style partitioning described above. This means that when
you call `open_dataset()` on this directory, you don't have to declare what the
partitions are because they can be read from the file paths.
(To instead write bare values for partition segments, i.e. `Cash` rather than 
`payment_type=Cash`, call `write_dataset()` with `hive_style = FALSE`.)

Perhaps, though, `payment_type == "Cash"` is the only data you ever care about,
and you just want to drop the rest and have a smaller working set.
For this, you can `filter()` them out when writing:

```r
ds %>%
  filter(payment_type == "Cash") %>%
  write_dataset("nyc-taxi/feather", format = "feather")
```

The other thing you can do when writing datasets is select a subset of columns 
or reorder them. Suppose you never care about `vendor_id`, and being a string column,
it can take up a lot of space when you read it in, so let's drop it:

```r
ds %>%
  group_by(payment_type) %>%
  select(-vendor_id) %>%
  write_dataset("nyc-taxi/feather", format = "feather")
```

Note that while you can select a subset of columns,
you cannot currently rename columns when writing a dataset.

## Partitioning performance considerations

Partitioning datasets has two aspects that affect performance: it increases the number of
files and it creates a directory structure around the files. Both of these have benefits
as well as costs. Depending on the configuration and the size of your dataset, the costs 
can outweigh the benefits. 

Because partitions split up the dataset into multiple files, partitioned datasets can be 
read and written with parallelism. However, each additional file adds a little overhead in 
processing for filesystem interaction. It also increases the overall dataset size since 
each file has some shared metadata. For example, each parquet file contains the schema and
group-level statistics. The number of partitions is a floor for the number of files. If 
you partition a dataset by date with a year of data, you will have at least 365 files. If 
you further partition by another dimension with 1,000 unique values, you will have up to 
365,000 files. This fine of partitioning often leads to small files that mostly consist of
metadata.

Partitioned datasets create nested folder structures, and those allow us to prune which 
files are loaded in a scan. However, this adds overhead to discovering files in the dataset,
as we'll need to recursively "list directory" to find the data files. Too fine
partitions can cause problems here: Partitioning a dataset by date for a years worth
of data will require 365 list calls to find all the files; adding another column with 
cardinality 1,000 will make that 365,365 calls.

The most optimal partitioning layout will depend on your data, access patterns, and which
systems will be reading the data. Most systems, including Arrow, should work across a 
range of file sizes and partitioning layouts, but there are extremes you should avoid. These
guidelines can help avoid some known worst cases:

 * Avoid files smaller than 20MB and larger than 2GB.
 * Avoid partitioning layouts with more than 10,000 distinct partitions.

For file formats that have a notion of groups within a file, such as Parquet, similar
guidelines apply. Row groups can provide parallelism when reading and allow data skipping
based on statistics, but very small groups can cause metadata to be a significant portion
of file size. Arrow's file writer provides sensible defaults for group sizing in most cases.

## Transactions / ACID guarantees

The dataset API offers no transaction support or any ACID guarantees.  This affects
both reading and writing.  Concurrent reads are fine.  Concurrent writes or writes
concurring with reads may have unexpected behavior.  Various approaches can be used
to avoid operating on the same files such as using a unique basename template for
each writer, a temporary directory for new files, or separate storage of the file
list instead of relying on directory discovery.

Unexpectedly killing the process while a write is in progress can leave the system
in an inconsistent state.  Write calls generally return as soon as the bytes to be
written have been completely delivered to the OS page cache.  Even though a write
operation has been completed it is possible for part of the file to be lost if
there is a sudden power loss immediately after the write call.

Most file formats have magic numbers which are written at the end.  This means a
partial file write can safely be detected and discarded.  The CSV file format does
not have any such concept and a partially written CSV file may be detected as valid.


