---
title: "Metadata"
description: > 
  Learn how Arrow uses Schemas to document structure of data objects, 
  and how R metadata are supported in Arrow
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metadata}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette describes the various data and metadata object types supplied by `arrow`, and documents how these objects are structured. 

```{r include=FALSE}
library(arrow, warn.conflicts = FALSE)
```

## Arrow metadata classes

The `arrow` package defines the following classes for representing metadata:

- A `Schema` is a list of `Field` objects used to describe the structure of a tabular data object; where
- A `Field` specifies a character string name and a `DataType`; and
- A `DataType` is an attribute controlling how values are represented


## R object attributes

Arrow supports custom key-value metadata attached to Schemas. When we convert a `data.frame` to an Arrow Table or RecordBatch, the package stores any `attributes()` attached to the columns of the `data.frame` in the Arrow object Schema. Attributes added to objects in this fasnion are stored under the `r` key, as shown below:

```{r}
# data frame with custom metadata
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
attr(df, "df_meta") <- "custom data frame metadata"
attr(df$y, "col_meta") <- "custom column metadata"

# when converted to a Table, the metadata is preserved
tb <- arrow_table(df)
tb$metadata
```

It is also possible to assign additional string metadata under any other key you wish, using a command like this:

```{r}
tb$metadata$new_key <- "new value"
```

Metadata attached to a Schema is preserved when writing the Table to Feather or Parquet. When reading those files into R, or when calling `as.data.frame()` on a Table or RecordBatch, the column attributes are restored to the columns of the resulting `data.frame`. This means that custom data types, including `haven::labelled`, `vctrs` annotations, and others, are preserved when doing a round-trip through Arrow.

Note that the attributes stored in `$metadata$r` are only understood by R. If you write a `data.frame` with `haven` columns to a Feather file and read that in Pandas, the `haven` metadata won't be recognized there. Similarly, Pandas writes its own custom metadata, which the R package does not consume. You are free, however, to define custom metadata conventions for your application and assign any (string) values you want to other metadata keys. For more details, see the documentation for `schema()`.


