---
title: "Reading and writing data files"
description: >
  Learn how to read and write CSV, Parquet, and Feather files with arrow 
output: rmarkdown::html_vignette
---

The arrow package provides functions for reading single data files into memory, in
several common formats. By default, calling any of these functions
returns an R data frame. To return an Arrow Table, set argument
`as_data_frame = FALSE`.

- `read_parquet()`: read a file in Parquet format
- `read_feather()`: read a file in the Apache Arrow IPC format (formerly called the Feather format)
- `read_delim_arrow()`: read a delimited text file (default delimiter is comma)
- `read_csv_arrow()`: read a comma-separated values (CSV) file
- `read_tsv_arrow()`: read a tab-separated values (TSV) file
- `read_json_arrow()`: read a JSON data file

For writing data to single files, the arrow package provides the
following functions, which can be used with both R data frames and 
Arrow Tables:

- `write_parquet()`: write a file in Parquet format
- `write_feather()`: write a file in Arrow IPC format
- `write_csv_arrow()`: write a file in CSV format

All these functions can read and write files in the local filesystem or
to cloud storage. For more on cloud storage support in arrow, see the [cloud storage article](./fs.html).

The arrow package also supports reading larger-than-memory single data files, and reading and writing multi-file data sets.
This enables analysis and processing of larger-than-memory data, and provides 
the ability to partition data into smaller chunks without loading the full 
data into memory. For more information on this topic, see the [dataset article](./dataset.html).

## Parquet format

[Apache Parquet](https://parquet.apache.org/) is a popular
choice for storing analytics data; it is a binary format that is 
optimized for reduced file sizes and fast read performance, especially 
for column-based access patterns. The simplest way to read and write
Parquet data using arrow is with the `read_parquet()` and 
`write_parquet()` functions. To illustrate this, we'll write the 
`starwars` data included in dplyr to a Parquet file, then read it 
back in. First load the arrow and dplyr packages:

```{r}
library(arrow, warn.conflicts = FALSE)
library(dplyr, warn.conflicts = FALSE)
```

Next we'll write the data frame to a Parquet file located at `file_path`:

```{r}
file_path <- tempfile()
write_parquet(starwars, file_path)
```

The size of a Parquet file is typically much smaller than the corresponding CSV 
file would have been. This is in part due to the use of file compression: by default, 
Parquet files written with the arrow package use [Snappy compression](https://google.github.io/snappy/) but other options such as gzip 
are also supported. See `help("write_parquet", package = "arrow")` for more
information.

Having written the Parquet file, we now can read it with `read_parquet()`:

```{r}
read_parquet(file_path)
```

The default is to return a data frame or tibble. If we want an Arrow Table instead, we would set `as_data_frame = FALSE`:

```{r}
read_parquet(file_path, as_data_frame = FALSE)
```

One useful feature of Parquet files is that they store data column-wise, and contain metadata that allow file readers to skip to the relevant sections of the file. That means it is possible to load only a subset of the columns without reading the complete file. The `col_select` argument to `read_parquet()` supports this functionality:

```{r}
read_parquet(file_path, col_select = c("name", "height", "mass"))
```

Fine-grained control over the Parquet reader is possible with the `props` argument. See `help("ParquetArrowReaderProperties", package = "arrow")` for details.

R object attributes are preserved when writing data to Parquet or
Arrow/Feather files and when reading those files back into R. This enables
round-trip writing and reading of `sf::sf` objects, R data frames with
with `haven::labelled` columns, and data frame with other custom
attributes. To learn more about how metadata are handled in arrow, the [metadata article](./metadata.html).

## Arrow/Feather format

The Arrow file format was developed to provide binary columnar 
serialization for data frames, to make reading and writing data frames 
efficient, and to make sharing data across data analysis languages easy.
This file format is sometimes referred to as Feather because it is an
outgrowth of the original [Feather](https://github.com/wesm/feather) project 
that has now been moved into the Arrow project itself. You can find the 
detailed specification of version 2 of the Arrow format -- officially 
referred to as [the Arrow IPC file format](https://arrow.apache.org/docs/format/Columnar.html#ipc-file-format) --
on the Arrow specification page. 

The `write_feather()` function writes version 2 Arrow/Feather files by default, and supports multiple kinds of file compression. Basic use is shown below:

```{r}
file_path <- tempfile()
write_feather(starwars, file_path)
```

The `read_feather()` function provides a familiar interface for reading feather files:

```{r}
read_feather(file_path)
```

Like the Parquet reader, this reader supports reading a only subset of columns, and can produce Arrow Table output:

```{r}
read_feather(
  file = file_path,
  col_select = c("name", "height", "mass"),
  as_data_frame = FALSE
)
```

## CSV format

The read/write capabilities of the arrow package also include support for 
CSV and other text-delimited files. The `read_csv_arrow()`, `read_tsv_arrow()`, 
and `read_delim_arrow()` functions all use the Arrow C++ CSV reader to read 
data files, where the Arrow C++ options have been mapped to arguments in a 
way that mirrors the conventions used in `readr::read_delim()`, with a 
`col_select` argument inspired by `vroom::vroom()`. 

A simple example of writing and reading a CSV file with arrow is shown below:

```{r}
file_path <- tempfile()
write_csv_arrow(mtcars, file_path)
read_csv_arrow(file_path, col_select = starts_with("d"))
```

In addition to the options provided by the readr-style arguments (`delim`, `quote`, `escape_doubple`, `escape_backslash`, etc), you can use the `schema` argument to specify column types: see `schema()` help for details. There is also the option of using `parse_options`, `convert_options`, and `read_options` to exercise fine-grained control over the arrow csv reader: see `help("CsvReadOptions", package = "arrow")` for details. 

## JSON format

The arrow package supports reading (but not writing) of tabular data from line-delimited JSON, using the `read_json_arrow()` function. A minimal example is shown below:

```{r}
file_path <- tempfile()
writeLines('
    { "hello": 3.5, "world": false, "yo": "thing" }
    { "hello": 3.25, "world": null }
    { "hello": 0.0, "world": true, "yo": null }
  ', file_path, useBytes = TRUE)
read_json_arrow(file_path)
```

## Further reading

- To learn more about cloud storage, see the [cloud storage article](./fs.html).
- To learn more about multi-file datasets, see the [datasets article](./dataset.html).
- The Apache Arrow R cookbook has chapters on [reading and writing single files](https://arrow.apache.org/cookbook/r/reading-and-writing-data---single-files.html) into memory and working with [multi-file datasets](https://arrow.apache.org/cookbook/r/reading-and-writing-data---multiple-files.html) stored on-disk.
