---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# arrow

[![conda-forge](https://img.shields.io/conda/vn/conda-forge/r-arrow.svg)](https://anaconda.org/conda-forge/r-arrow)

R integration with Apache Arrow.

## Installation

You first need to install the C++ library:

### macOS

On macOS, you may install using homebrew:

```
brew install apache-arrow
```

### From source

First install a release build of the C++ bindings to arrow.

```shell
git clone https://github.com/apache/arrow.git
cd arrow/cpp && mkdir release && cd release

# It is important to statically link to boost libraries
cmake .. -DARROW_PARQUET=ON -DCMAKE_BUILD_TYPE=Release -DARROW_BOOST_USE_SHARED:BOOL=Off
make install
```

## Then the R package

### From github (defaults to master branch)

```r
library("devtools")
devtools::install_github("apache/arrow/r")
```

## Example

```{r}
library(arrow)

(tib <- tibble::tibble(x = 1:10, y = rnorm(10)))
tab <- table(tib)
tab$schema
tab
as_tibble(tab)
```

## Developing

The arrow package is using devtools for package related manipulations. To install:

```r
library.packages("devtools")
```

Within a local clone, one can test changes by running the test suite.

### Running the test suite

```r
library("devtools")
devtools::install_dev_deps()
devtools::test()
```

### Full package validation

```shell
R CMD build --keep-empty-dirs .
R CMD check arrow_*.tar.gz --as-cran --no-manual
```
