---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# rrrow

rrrow is an R front end to Apache Arrow. 

## Installation

I've only tested this locally for now, here is how I install it, borrowed
from [rarrow](https://github.com/jimhester/rarrow). 

First you need the Arrow C++ library installed

```
git clone https://github.com/apache/arrow.git
cd arrow/cpp && mkdir release && cd release

# It is important to statically link to boost libraries
cmake .. -DCMAKE_BUILD_TYPE=Release -DARROW_BOOST_USE_SHARED:BOOL=Off
make install
```

Then you can install the development version of the package: 

```r
# devtools::install_github("apache/arrow/r")
devtools::install_github("romainfrancois/arrow/r")
```


## Example

factory functions: 

```{r}
library(rrrow)

# metadata factories
int32()
struct( x = int32(), y = int64() )
```

ArrayBuilder

```{r}
# make a builder for an array of type int32. 
(b <- ArrayBuilder( int32() ) )
ArrayBuilder__num_children(b)
```

This is just kicking the tires with the C++ class system for now. Eventully, we
probably will have something like `b$num_children()` or perhaps the `ArrayBuilder` 
class won't even be public. 

At the moment the internal code is hand written, we are considering moving to 
[RcppR6](https://github.com/richfitz/RcppR6) as a development time dependency to 
generate the bindings. 
